---
title: "Stat 324 HW7: Hypothesis Tests for 2 Independent and Dependent Populations"
author: "Devin Johnson, Dis. 313"
date: "November 13, 2018"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)
```


## Problem 1
```{r}
# Data for problem 1
d2 = c(422,298,423,346,375,431,290)
d1 = c(368,293,401,314,384,404,267)
diffs = c()
for(i in 0:length(d1)){
  diffs = append(diffs, d2[i] - d1[i])
}
```

**--1A: Hypothesis Test--**  
$H_o:\mu_{d2} - \mu_{d1} = 5$  
$H_a:\mu_{d2} - \mu_{d1} > 5$  
$\alpha = 0.10$

```{r}
# Calculate matched pair test
tobs = (mean(diffs) - 5) / (sd(diffs)/sqrt(length(diffs)))
pval = pt(q = tobs, df = length(diffs) - 1, lower.tail = FALSE)
cat(paste(" T stat:", tobs, "\n", "pval:", pval), "\n")

# Check w/ R
t.test(d2, d1, alternative = "greater", paired = TRUE, mu=5)
```

Since our p value < $\alpha$, we reject the null.

**--1B: Assumptions for Part A--**

* Populations are not independent  
* IID samples  
* Population of differences is normal or large enough sample size

**--1C: 90% Confidence Interval--**
```{r}
# Calculate CI
l = mean(diffs) - (abs(qt(0.10/2, length(diffs) - 1)) * (sd(diffs)/sqrt(length(diffs))))
u = mean(diffs) + (abs(qt(0.10/2, length(diffs) - 1)) * (sd(diffs)/sqrt(length(diffs))))
cat(paste(" Lower", l, "\n", "Upper:", u), "\n")

# Check w/ R (using two sided test for two sided CI)
t.test(d2, d1, alternative = "two.sided", paired = TRUE, mu=5, conf.level = 0.90)
```

The confidence interval tells us that we are 90% confident that the true difference in means falls within the given interal while the hypothesis only gives us evidence that the difference is > 5.

## Problem 2
 * Wilcoxon Rank Sum p-value: **0.028** (does not change since ranks are the same)  
 * T test p-value: **Smaller than 0.006**. (the t-test will be more affected by larger differences in the two samples since it is based off real values and not just rank. This leads to more evidence and thus a smaller p value.)

## Problem 3
A paired design would mean that both samples are from the same test subjects. Since both measurments are taken after the same rainfall dates, one could argue that they are not necessarily independent. Thus we say paired.

In this question we are not concerned with assumptions made so I choose:

* Wilcoxon Signed Rank
* Paired sample t test
* Wilcoxon Signed rank test
* Bootstrap test for one sample (differences)

## Problem 4
```{r}
# Data for problem 4
warm = c(214, 194, 221, 198, 212)
cold = c(225, 215, 253, 272, 254)
```
$H_o:\mu_{cold} - \mu_{warm} = 0$  
$H_a:\mu_{cold} - \mu_{warm} > 0$  
$\alpha = 0.10$

**--4A: Using Normal Assumption--**  
We can use the 2 sample t test with either equal variance or non equal variance assumptions.  
```{r}
t.test(cold, warm, alternative = "greater", var.equal = FALSE)
t.test(cold, warm, alternative = "greater", var.equal = TRUE)
```

**--4B/C: Using Non-Normal Assumption--**  
We can use the bootstrap, wilcoxon rank sum, or permutation tests under the assumption that both populations independent and independent samples were taken.
```{r}
# Bootstrap function
boottwo = function(dat1, dat2, nboot) {
  bootstat = numeric(nboot)
  obsdiff = mean(dat1) - mean(dat2)
  n1 = length(dat1)
  n2 = length(dat2)
  for(i in 1:nboot) {
    samp1 = sample(dat1, size = n1, replace = T)
    samp2 = sample(dat2, size = n2, replace = T)
    bootmean1 = mean(samp1)
    bootmean2 = mean(samp2)
    bootvar1 = var(samp1)
    bootvar2 = var(samp2)
    bootstat[i] = ((bootmean1 - bootmean2) - obsdiff)/sqrt((bootvar1/n1) + (bootvar2/n2))
}
  return(bootstat)
}

# Perform bootstrap
set.seed(1)
boot = boottwo(cold, warm, 10000)
t_obs = (mean(cold) - mean(warm)) / sqrt((var(cold) / length(cold))  + (var(warm) / var(warm)))
m_high = sum(boot > t_obs)
pval = m_high/ 10000
cat(paste(" T observed:", t_obs, "\n", "P Value:", pval))

# Wilcoxon Rank Sum
wilcox.test(cold, warm, alternative = "greater")

# Permutation function
permtwo <- function(dat1, dat2, nperm) {
	permstat <- NULL
	for(i in 1:nperm) {
		n1 <- length(dat1)  #find length of first sample
		n2 <- length(dat2)  #find length of second sample
		alldat <- c(dat1, dat2)  #combine into 1 large sample
		samp <- sample(alldat, replace=FALSE)  #shuffle the combined vector of values
		pmean1 <- mean(samp[1:n1])   #find mean of first n1 observations as x1*-bar
		pmean2 <- mean(samp[(n1+1):(n1+n2)]) #find mean of last n2 observations as x2*-bar
		permstat[i] <- pmean1 - pmean2
}
	return(permstat)
}

# Perform the permutation
set.seed(1)
perm = permtwo(cold, warm, 10000)
t_obs = mean(cold) - mean(warm)
m_high = sum(perm >= t_obs)
pval = m_high/10000
cat(paste(" T observed:", t_obs, "\n", "P Value:", pval))
```

**--4D: Test Conclusions--**
P values are below $\alpha$, so we hav sufficient evidence to reject the null hypothesis. This would suggest to the the scientists that cold temperatures could indeed raise blood pressure.


## Problem 5
```{r}
# Data for problem 5
n_urb = 140
n_suburb = 160
n_rural = 200
success_urb = 83
success_suburb = 92
success_rural = 86
```

**--5A: Hypothesis Test--**  
$H_o: \pi_{urb} - \pi_{suburb} = 0$  
$H_a: \pi_{urb} - \pi_{suburb} \neq 0$  
$\alpha = 0.05$  

Asssumptions:

* Populations independent
* IID samples
* Popualations approximately normal (by CLT)

```{r}
pihat_urb = success_urb / n_urb
pihat_suburb = success_suburb / n_suburb
pihat_rural = success_rural / n_rural
pihat_ha = (success_suburb + success_urb) / (n_urb + n_suburb)

z_obs = (pihat_urb - pihat_suburb - 0) / sqrt(pihat_ha * (1-pihat_ha) * ((1/n_urb) + (1/n_suburb)))
pval = 2*pnorm(-abs(z_obs))
cat(paste(" Z observed:", z_obs, "\n", "P Value:", pval))
```

Fail to reject null since pval $<\alpha$

**--5B: 95% Confidence Interval--**  
Asssumptions:

* Populations independent
* IID samples
* Popualations approximately normal (by CLT)

```{r}
print(pihat_urb - pihat_rural)
l = (pihat_rural - pihat_urb) - (1.96*sqrt((pihat_urb * (1-pihat_urb) / n_urb) + (pihat_rural * (1-pihat_rural) / n_rural)))
u = (pihat_rural - pihat_urb) + (1.96*sqrt((pihat_urb * (1-pihat_urb) / n_urb) + (pihat_rural * (1-pihat_rural) / n_rural)))
cat(paste(" Lower", l, "\n", "Upper:", u), "\n")
```

